On The Impact of Atoms of Confusion in JavaScript Code

### 2022-03-29 3_Method.tex
First version with the combined methodology section. The initial part is written starting from the assumption that the MSR section will be moved. We can review this later. I still need to add a big table with all the atom candidates and make a complete pass to avoid duplication and inconsistencies. I did a first pass and addressed some of that already.

\begin{enumerate}[(RQ1)]
\item \rqa 
\item \rqb
% \item \rqc  
\item \rqd
\end{enumerate}

Answering these research questions has several implications. For instance, we can either generalize or refute the perceptions about atoms of confusion already discussed in the literature (goal of research questions (RQ1) and (RQ2)). In addition, answers to the third and fourth questions allow us to enrich existing catalogs about atoms of confusion and discuss how often they occur in practice. Answering these research questions also lays the foundations for the implementation of tools that can automatically transform code into cleaner versions.
%---though we postpone these results to future research work.
We conduct a mixed-methods study to answer these questions, including a survey, a set of interviews, and a study for mining JavaScript software repositories.

\subsection{Survey Study Setting}\label{sec:meth:survey}

In the first study, we explore the impact of atom candidates on
understanding JavaScript code (goal of research question RQ1).	
To this end, we investigate whether or not code snippets that
contain atom candidates (a) produce a higher
misunderstanding rate when programmers
try to predict their outcome and (b) require more time for programmers to predict their output.
Since JavaScript and \clang have some constructs in common (Section~\ref{back}), we first selected a set of 9+1 atom candidates: nine previously-validated atoms of confusion for the \clang language~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} that also exist in JavaScript programs plus one atom candidate that is specific to the JavaScript language (Automatic Semicolon Insertion).
Our supplementary material discusses and shows examples of the atoms we consider in our research.

\subsubsection*{Survey design}

The design of our first study blocks two variables
(developer experience and the code snippets) and
considers two treatments: the presence or absence of 
atoms candidates within the code snippets. 
To achieve such a design goal, controlling the effect of experience and individual code snippets, we resorted to the \textit{Latin Square Design} \cite{Hunter-Experimenters}. Using this design we create a 2 x 2 matrix in which each row represents a subject and each column indicates the set of code snippets. The design of each square (a replica) is such that no treatment is repeated in the same row or column. For example, considering that we have a set of 10
code samples $s_0, s_1, ..., s_{10}$, if a given subject (P1) is asked to predict the output of the code samples $s_0, s_1, ..., s_5$ that contain atom candidates, then, when answering questions about non-confusing code snippets, they will only be presented with non-confusing versions of the code samples $s_6, s_7,..., s_{10}$. Furthermore, a given subject (P2), who constitutes the second row of our example square, will be asked questions about the non-confusing versions for $s_1, s_2, ..., s_5$, and will answer questions about confusing snippets for $s_6, s_7,..., s_{10}$. By doing that, we guarantee that all versions of the code snippets are contained within each square, and that each configuration occurs only once within a square. Figure \ref{fig:latinsquare} offers a visual representation of the concept.

@@

\subsubsection*{Survey Instrument}

We implemented our survey as a web application. As part of this effort, we carried out an informal pilot whose main objectives were (i) to spot bugs in the application and in the data collection mechanism; (ii) to gain feedback from respondents about the user experience of the application; and (iii) to formulate an estimate about how long answering the survey would take on average. Undergraduate students and professional colleagues took the pilot survey. Some users reported layout defects, and many reported that the landing page did not explain the survey well enough. We also spotted minor issues with our routines to create and populate the Latin Squares.

@@

The last section of the survey presented a sequence of ten questions, each containing a code snippet. For each question, there was a text box where the answer should be written. There was also an ``I do not know'' button, which, when clicked, led the subject to the next question. In our setting, ``I do not know'' was treated as a wrong answer. The code snippets were presented as images copied from a text editor, so as to demotivate respondents from resorting to external resources by copying and pasting the code into an interpreter. Upon submitting their answer for a particular question, the subject was automatically led to a similar page, containing another snippet.

We decided not to provide feedback about the time students took to answer each question. We also did not tell them whether their answers were correct or not. Our main concern was to avoid introducing bias for future respondents. Since we posted the survey on a social media platform, possible threats could have arisen if we gave respondents instant feedback.

@@

\subsubsection*{Survey audience}

>>>>

\begin{enumerate}[(RQ1)]
\item \rqa 
\item \rqb
% \item \rqc  
\item \rqd\castor{This one will probably be removed.}
\end{enumerate}

By answering these research questions, we can either generalize or refute the findings about atoms of confusion already discussed in the literature (goal of research questions (RQ1) and (RQ2)). In addition, answer to the third question allows us to enrich existing catalogs about atoms of confusion and discuss how often they occur in practice\castor{This should probably be removed.}. Answering these research questions also lays the foundations for the implementation of tools that can automatically transform code into cleaner versions.
%---though we postpone these results to future research work.
We conduct a mixed-methods study to answer these questions. It includes two independently designed and conducted experiments\castor{I would rather use the term ``experiment''.}: a replication of the first study on atoms of confusion, but targeting JavaScript programs, and a new study using a latin-square counterbalancing experimental design to control for subject experience. 
These two experiments explore the impact of atom candidates on understanding JavaScript code (goal of research question RQ1). They investigate whether or not code snippets that contain atom candidates produce a higher misunderstanding rate when programmers try to predict their outcome.\castor{I've removed the part about time because, in order for the first study to be a replication, we need to compute time differently, accounting for incorrect responses as well.}% and (b) require more time for programmers to predict their output. 
We leverage two independently-developed studies with different designs to more rigorously verify the impact of atoms of confusion in code comprehension activities. 
We also conduct a set of interviews with developers to contrast the quantitative results of the experiments with their preferences and opinions (thus addressing RQ2).

\subsection{Replication Study Setting}\label{sec:meth:replication}

We call the first study we performed the \textit{``Replication Study''} because it is a partial replication~\cite{Juristo2012} of the first study on atoms of confusion, conducted by Gopstein et. al~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17}, but targeting JavaScript instead of the C language. Gopstein et. al. employed a randomized, partial counterbalancing experimental design on a sample of 73 subjects, mostly students, with three or more months of experience in C or C++. The control variable consisted of a tiny program containing a single atom candidate and the treatment a functionally-equivalent version that does not contain the atom. The main dependent variable is whether subjects are able to correctly determine the output of the programs. The other dependent variable is the time required to correctly vs. incorrectly determine the out of the programs. The study had the null hypothesis that \textit{``code from both control and treatment groups can be hand-evaluated with equal accuracy''} and the alternate hypothesis that \textit{``the existence of an atom of confusion causes more errors than other code in hand-evaluated outputs.''}. For each atom candidate, six programs were built, three versions including the atom candidate and three versions not including it. Each subject was exposed to four programs for every atom candidate, two including an atom and two not including an atom.

\subsubsection*{Experimental design}

In this study we adopt the same experimental design, control and treatment variables, dependent variable, and hypotheses. We have also built six programs for each atom candidate, three with and three without the atom. Considering the elements that can vary in a replication study, defined by Juristo and Gomez~\cite{Juristo2012}, in this study we employ the same experimental protocol and construct operationalizations. We have different population properties, experimenters, and a different site. In addition, since we focus on a different language, some atoms that exist in C cannot by construction exist in JavaScript programs. Conversely, JavaScript exhibits atom candidates that do not exist in C.

The design of the replication study considers two treatments: the presence or absence of atoms candidates within the programs. To identify atom candidates, we start out by selecting every atom candidate from Gopstein's study that can also happen in JavaScript. That excludes atoms related to pointers and preprocessor directives. Furthermore, we conduct an informal search of programming forums, solutions to Code Golf~\footnote{https://codegolf.stackexchange.com/} problems, code style guidelines, and expert knowledge to identify new atom candidates. Overall, the replication study analyzes 24 atoms, 15 from the original study and 9 that are new and specific to JavaScript. Table~\ref{tab:atoms_replication} presents the complete list.

For each atom candidate, we write six short programs, three including and three not including the atom candidate. The former are the control for the study or the obfuscated versions of the programs. The alternative, atom-free, versions of the programs are the treatment, or clean, versions. Each subject is exposed to one obfuscated and one clean version of each atom, totalizing 48 programs, and should determine their outputs. As mentioned before, we measure answer correctness and the time to answer questions correctly vs. incorrectly. We control for learning effects~\cite{Neely:1991:SPE} in three manners. First, by having multiple obfuscated and clean versions for each atom and presenting only one of each per subject. Second, by presenting the programs in a random order. Third, by presenting obfuscated and clean versions corresponding to the same atom candidate with at least 11 other programs in between, in  accordance to the original experimental protocol~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17}.

\subsubsection*{Study instrument}

The study is conducted by means of a web application. As part of this effort, we carried out an informal pilot whose main objectives were (i) to spot bugs in the application and in the data collection mechanism; (ii) to gain feedback from respondents about the user experience of the application; and (iii) to formulate an estimate about how long answering the survey would take on average. Based on the feedback of the participants of the pilot, we decided to present only one obfuscated and one clean version of each atom candidate, to reduce to total time of the study and potentially increase participation. Initially, our plan was to present two of each.

We organize the study in three sections. In the first one we present instructions and also a check button whose checking meant users agreed that all collected data would be used anonymously and solely for research purposes. The instructions explain how the survey works and asks them to dedicate their attention to it.  We stressed to participants the importance of not using any aids during the survey, such as online or console interpreters.

The second section presents the programs, one at a time. For each one, there is a text box where the answer should be written. There is also an ``I do not know'' button, which, when clicked, leads the subject to the next question. In our setting, ``I do not know'' was treated as a wrong answer. To reduce the likelihood of a subject leveraging attempting to execute the programs, the web application verifies whether the subject attempts to change tabs or windows and presents the programs as images, instead of text. Upon submitting an answer for a particular program, the subject is automatically led to a similar page, containing the next one.

We do not provide feedback about the time subjects take to answer each question. We also do not tell them whether their answers are correct or not. This aims at avoiding introducing bias for future respondents. Since we posted the survey on social media platforms, possible threats could have arisen if we gave respondents instant feedback.

\subsubsection*{Study audience}

We posted the survey on the authors' social media platforms, on Brazilian CS departments' mailing lists, and on two programming subreddits. We explained our research purposes and asked developers to take the survey only if they have some familiarity with JavaScript. We collected 70 responses. The average survey respondent has a mean 26 months of experience in JavaScript, with a median of 12.

\subsection{Latin-Square Study Setting}\label{sec:meth:survey}

This experiment complements the replication study by using a different experimental design and controlling for one additional variable, the experience of the subjects. For this study we employ a subset of the atom candidates used in the replication study. More specifically, we
%Since JavaScript and \clang have some constructs in common (Section~\ref{back}), we first 
select a set of 9+1 atom candidates: nine previously-validated atoms of confusion for the \clang language~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} that also exist in JavaScript programs plus one atom candidate that is specific to the JavaScript language (Automatic Semicolon Insertion)\castor{I think the name of this atom is inconsistent between the studies}. 
Our supplementary material discusses and shows examples of the atoms we consider in our research.\castor{We should probably do this here or just refer to the table that I intend to insert for the replication study.}

\subsubsection*{Experimental design}

The design of our second study blocks two variables
(developer experience and the code snippets) and
considers two treatments: the presence or absence of 
atoms candidates within the code snippets. 
To achieve such a design goal, controlling the effect of experience and individual code snippets, we resorted to the \textit{Latin Square Design} \cite{Hunter-Experimenters}. Using this design we create a 2 x 2 matrix in which each row represents a subject and each column indicates the set of code snippets. The design of each square (a replica) is such that no treatment is repeated in the same row or column. For example, considering that we have a set of 10
code samples $s_0, s_1, ..., s_{10}$, if a given subject (P1) is asked to predict the output of the code samples $s_0, s_1, ..., s_5$ that contain atom candidates, then, when answering questions about non-confusing code snippets, they will only be presented with non-confusing versions of the code samples $s_6, s_7,..., s_{10}$. Furthermore, a given subject (P2), who constitutes the second row of our example square, will be asked questions about the non-confusing versions for $s_1, s_2, ..., s_5$, and will answer questions about confusing snippets for $s_6, s_7,..., s_{10}$. By doing that, we guarantee that all versions of the code snippets are contained within each square, and that each configuration occurs only once within a square. Figure \ref{fig:latinsquare} offers a visual representation of the concept.

@@

\subsubsection*{Study instrument}

We implemented our survey as a web application and carried out a pilot, similarly to the replication study. 
%As part of this effort, we carried out an informal pilot whose main objectives were (i) to spot bugs in the application and in the data collection mechanism; (ii) to gain feedback from respondents about the user experience of the application; and (iii) to formulate an estimate about how long answering the survey would take on average.
Undergraduate students and professional colleagues took the pilot survey. Some users reported layout defects, and many reported that the landing page did not explain the survey well enough. We also spotted minor issues with our routines to create and populate the Latin Squares.

@@

The last section of the survey presented a sequence of ten questions, each containing a code snippet. For each question, there was a text box where the answer should be written. There was also an ``I do not know'' button, which, when clicked, led the subject to the next question. In our setting, ``I do not know'' was treated as a wrong answer. The code snippets were presented as images copied from a text editor, so as to demotivate respondents from resorting to external resources by copying and pasting the code into an interpreter. Upon submitting their answer for a particular question, the subject was automatically led to a similar page, containing another snippet. Similarly to the replication study, we do not provide feedback about time and correctness to the subjects.

@@

\subsubsection*{Study audience}
